---
title: "Why is my OS X Yosemite Install Taking So Long?: An Analysis"
author: "Tony Fischetti"
date: "October 21, 2014"
output: html_document
---

-------

### why?
Since the latest Mac OS X update, 10.10 "Yosemite", was released last Thursday,
there have been complaints springing up online of the progress bar
woefully underestimating the actual time to complete installation. More
specifically, it appeared as if, for a certain group of people (myself included),
the installer would stall out at "two minutes remaining" or "less than a minute
remaining"-sometimes for hours.

In the vast majority of these cases, though, the installation process didn't
hang, it was just performing a bunch of unexpected tasks that it couldn't
predict.

During the install, striking "Command" + "L" would bring up the install logs. 
In my case, the logs indicated that the installer was busy right until the very
last minute.


(a) Not knowing very much about OS X's installation process and wanting to learn
more, and
(b) wanting to answer why the installation was taking longer that the progress bar expected, I saved the log to a file on my disk with the intention of analyzing
it before the installer automatically restarted my computer.


### cleaning

The log file from the Yosemite installer wasn't in a format that R (or any
program) could handle natively so before we can use it, we have to clean/munge
it. To do this, we'll write a program in the queen of all text-processing
languages: perl.

This script will read the log file, line-by-line from standard input (for
easy shell piping), and spit out nicely formatted tab-delimited lines.

```
#!/usr/bin/perl

use strict;
use warnings;

# read from stdin
while(<>){
    chomp;
    my $line = $_;
    my ($not_message, $message) = split ': ', $line, 2;

    # skip lines with blank messages
    next if $message =~ m/^\s*$/;

    # get rid of bracketed numbers
    $not_message =~ s/\[\d+\]//;

    my ($month, $day, $time, $machine, $service) = split " ", $not_message;

    print join("\t", $month, $day, $time, $machine, $service, $message) . "\n";
}
```

We can output the cleaned log file with this shell command:

```
echo "Month\tDay\tTime\tMachine\tService\tMessage" > cleaned.log
grep '^Oct' ./YosemiteInstall.log | grep -v ']:  ' | grep -v ': }' |  ./clean-log.pl >> cleaned.log
```

This cleaned log contains 6 fields: 'Month', 'Day', 'Time', 'Machine (host)', 'Service', and 'Message'. The installation didn't span days (it didn't even
span an hour) so *technically* I didn't need the 'Month' and 'Day' fields, but
I left them in for completeness' sake.

### analysis

Let's set some options and load the libraries we are going to use:

```{r, message=FALSE, warning=FALSE}
# options
options(echo=TRUE)
options(stringsAsFactors=FALSE)

# libraries
library(dplyr)
library(ggplot2)
library(lubridate)
library(reshape2)
```


Let's read the log file that I cleaned and add a few columns with correctly
parsed timestamp using lubridate's "parse_date_time()" function

```{r, warning=FALSE}
yos.log <- read.delim("./cleaned.log", sep="\t") %>%
  mutate(nice.date=paste(Month, Day, "2014", Time)) %>%
  mutate(lub.time=parse_date_time(nice.date, 
                                  "%b %d! %Y! %H!:%M!:%S!", 
                                  tz="EST"))
```

And remove the rows of dates that didn't parse correctly
```{r}
yos.log <- yos.log[!is.na(yos.log$lub.time),]

head(yos.log)
```


The first question I had was how long the installation process took

```{r}
install.time <- yos.log[nrow(yos.log), "lub.time"] - yos.log[1, "lub.time"]
(as.duration(install.time))
```

Ok, about a half-hour.

Let's make a column for cumulative time by subtracting each row's time by
the start time
```{r}
yos.log$cumulative <- yos.log$lub.time - min(yos.log$lub.time, na.rm=TRUE)
```


In order to see what processes were taking the longest, we have to make a
column for elapsed time. To do this, we can subtract each row's time from the
time of the subsequent row.

```{r}
yos.log$elapsed <- lead(yos.log$lub.time) - yos.log$lub.time

# remove last row
yos.log <- yos.log[-nrow(yos.log),]
```


Which services were responsible for the most writes to the log and what services
took the longest? We can find out with the following elegant dplyr construct.
While we're at it, we should add columns for percentange of the whole for
easy plotting

```{r}
counts <- yos.log %>%
  group_by(Service) %>%
  summarise(n=n(), totalTime=sum(elapsed)) %>%
  arrange(desc(n)) %>%
  top_n(8, n) %>%
  mutate(percent.n = n/sum(n)) %>%
  mutate(percent.totalTime = as.numeric(totalTime)/sum(as.numeric(totalTime)))
(counts)
```

Ok, the "OSInstaller" is responsible for the vast majority of the writes to the
log and to the total time of the installation. "opendirectoryd" was the next
most verbose process, but its processes were relatively quick compared to the
"Unknown" process' as evidenced by "Unknown" taking almost 4 times longer, in
aggregate, in spite of having only 7% of "opendirectoryd"'s log entries.

We can more intuitively view the number-of-entries/time-taken mismatch thusly:
```{r, message=FALSE}
melted <- melt(as.data.frame(counts[,c("Service",
                                       "percent.n",
                                       "percent.totalTime")]))

ggplot(melted, aes(x=Service, y=as.numeric(value), fill=factor(variable))) +
  geom_bar(width=.8, stat="identity", position = "dodge",) +
  ggtitle("Breakdown of services during installation by writes to log") +
  ylab("percent") + xlab("service") +
  scale_fill_discrete(name="Percent of",
                      breaks=c("percent.n", "percent.totalTime"),
                      labels=c("writes to logfile", "time elapsed"))
```

As you can see, the "Unknown" process took a disproportionately long time
for its relatively few log entries; the opposite behavior is observed with 
"opendirectoryd". The other processes contribute very little to both the number
of log entries and the total time in the installation process.

What were the 5 most lengthy processes?
```{r}
yos.log %>%
  arrange(desc(elapsed)) %>%
  select(Service, Message, elapsed) %>%
  head(n=5)
```

The top processes were:

1. Unpacking and moving the contents of "Essentials.pkg" into what is to become
   the newsystem directory structure. This ostensibly contains items like all
   the updated applications (Safari, Mail, etc..). **(almost three minutes)**
   
2. Archiving the old system logs **(two and a half minutes)**

3. Loading the kernel module that allows the onboard serial ATA controller
   to work **(a little over a minute)**
  

Let's view a density plot of the number of writes to the log file during
installation.
```{r}
ggplot(yos.log, aes(x=lub.time)) +
  geom_density(adjust=3, fill="#0072B2") +
  ggtitle("Density plot of number of writes to log file during installation") +
  xlab("time") + ylab("")
```

This graph is very illuminating; the vast majority of log file writes were 
the result of very quick processes that took place in the last 15 minutes of
the install, **which is when the progress bar read that only two minutes
were remaining.**

In particular, there were a very large number of log file writes between 11:47
and 11:48; what was going on here?

```{r}

# if the first time is in between the second two, this returns TRUE
is.in <- function(time, start, end){
  if(time > start && time < end)
    return(TRUE)
  return(FALSE)
}

the.start <- ymd_hms("14-10-18 11:47:00", tz="EST")
the.end <- ymd_hms("14-10-18 11:48:00", tz="EST")

# logical vector containing indices of writes in time interval
is.in.interval <- sapply(yos.log$lub.time, is.in,
                         the.start,
                         the.end)

# extract only these rows
in.interval <- yos.log[is.in.interval, ]

# what do they look like?
silence <- in.interval %>%
  select(Message) %>%
  sample_n(7) %>%
  apply(1, function (x){cat("\n");cat(x);cat("\n")})
```

Ah, so these processes are the result of the installer having to move files
back into the new installation directory structure. In particular, the vast
majority of these move operations are moving files related to a program called
"texlive". I'll explain why this is to blame for the inaccurate projected time
to completion in the next section.


But lastly, let's view a faceted density plot of the number of log files writes by
process. This might give us a sense of what steps go on as the installation
progresses by showing us with processes are most active.

```{r}
# reduce number of service to a select few of the most active
smaller <- yos.log %>%
  filter(Service %in% c("OSInstaller", "opendirectoryd",
                        "Unknown", "OS"))

ggplot(smaller, aes(x=lub.time, color=Service)) +
  geom_density(aes( y = ..scaled..)) +
  ggtitle("Faceted density of log file writes by process (scaled)") +
  xlab("time") + ylab("")
```


This shows that no one process runs consistently throughout the entire
installation process, but rather that the process run in spurts.


### the answer

The vast majority of Mac users don't place strange files in certain special
system-critical locations like '/usr/local/' and '/Library/'. Among those
who do, though, these directories are littered with hundreds and hundreds
of custom files that the installer doesn't and can't have prior knowledge of.

In my case, and probably many others, the estimated time-to-completion was
inaccurate because the installer couldn't anticipate needing to copy
back so many files to certain special directories after unpacking
the contents of the new OS. Additionally, for each of these copied files,
the installer had to make sure the subdirectories had the exact same meta-data
(permissions, owner, reference count, creation date, etc...) as before the
installation began. This entire process added many minutes to the procedure
at a point when the installer thought it was pretty much done.



What were some of the files that the installer needed to copy back?
The answer will be different for each system but, as mentioned above, anything
placed '/usr/local' and '/Library' directories that wasn't Apple-supplied
needed to be moved and moved back.

**/usr/local/**
/usr/local/ is used chiefly for user-installed software that isn't part
of the OS distribution. In my case, my /usr/local contained a custom compliled
Vim; ClamXAV, a lightweight virus scanner that I use only for the benefit of my
Windows-using friends; and texlive, software for the TeX typesetting system. 
texlive was, by far, the biggest time-sink since it had over 123,491 files.

In addition to these programs, many users might find that the Homebrew package
manager is to blame for their long installation process, since this software
also uses the /usr/local prefix (although it probably should not).

**/Library/**
Among other things, this directory holds (subdirectories that hold) modules
and packages that the Apple-supplied Python, Ruby, and Perl uses. If you
use these Apple-supplied versions of these languages and you install your
own packages/modules using super-user privileges, the new packages will go
into this directory and it will appear foreign to the Yosemite installer.

To get around this issue, either install packages/modules in a local
(non-system) library, or use alternate versions of these programming languages
that you either download and install yourself, or use MacPorts to install.
