---
title: "Why is my OS X Yosemite Install Taking So Long?: An Analysis"
author: "Tony Fischetti"
date: "October 21, 2014"
output: html_document
---

-------

### why?
Since the latest Mac OS X update, 10.10 "Yosemite", was released last Thursday,
there have been complaints springing up online 


### cleaning

The log file from the Yosemite installer wasn't in a format that R (or any
program) could handle natively so before we can use it, we have to clean/munge
it. To do this, we'll write a program in the mother of all text-processing
languages: perl.

This script will read the log file, line-by-line from standard input (for
easy shell piping), and spit out nicely formatted tab-delimited lines.

```
#!/usr/bin/perl

use strict;
use warnings;

# read from stdin
while(<>){
    chomp;
    my $line = $_;
    my ($not_message, $message) = split ': ', $line, 2;

    # skip lines with blank messages
    next if $message =~ m/^\s*$/;

    # get rid of bracketed numbers
    $not_message =~ s/\[\d+\]//;

    my ($month, $day, $time, $machine, $service) = split " ", $not_message;

    # print "Month: $month\n";
    # print "Day: $day\n";
    # print "Time: $time\n";
    # print "Machine: $machine\n";
    # print "Service: $service\n\n";

    print join("\t", $month, $day, $time, $machine, $service, $message) . "\n";
}
```

We can output the cleaned log file with this shell command:

```
echo "Month\tDay\tTime\tMachine\tService\tMessage" > cleaned.log
grep '^Oct' ./YosemiteInstall.log | grep -v ']:  ' | grep -v ': }' |  ./clean-log.pl >> cleaned.log
```

### analysis

Let's set some options and load the libraries we are going to use:

```{r, message=FALSE, warning=FALSE}
# options
options(echo=TRUE)
options(stringsAsFactors=FALSE)

# libraries
library(dplyr)
library(ggplot2)
library(lubridate)
library(reshape2)
```


Let's read the log file that I cleaned and add a few columns with correctly
parsed timestamp

```{r, warning=FALSE}
yos.log <- read.delim("./cleaned.log", sep="\t") %>%
  mutate(nice.date=paste(Month, Day, "2014", Time)) %>%
  mutate(lub.time=parse_date_time(nice.date, 
                                  "%b %d! %Y! %H!:%M!:%S!", 
                                  tz="EST"))
```

And remove the rows of dates that didn't parse correctly
```{r}
yos.log <- yos.log[!is.na(yos.log$lub.time),]

head(yos.log)
```


